---
title: "VectorByte Methods Training: Introduction to Time Series and Forecasting (practical)"
author:
  - name: 
      given: Alicia
      family: Arneson
    affiliation: Virginia Tech and VectorByte
  - name: 
      given: Leah R.
      family: Johnson
    url: https://lrjohnson0.github.io/QEDLab/leahJ.html
    affiliation: Virginia Tech and VectorByte
citation: true
date: 2025-06-01
date-format: long
format:
  html:
    toc: true
    toc-location: left
    html-math-method: katex
    css: styles.css
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

set.seed(123)
```

# Overview and Instructions

The goals of this practical are to:

1.  Practice setting up data for forecasting

2.  Practice fitting time series models for forecasting

3.  Practice generating forecasts for a given model and evaluating its performance

4.  Learn aggressively.... See the forecasting challenge section below :)

This practical will take you through an example of modeling a terrestrial phenology data set (how much greenness can be observed from above) from NEON and generating forecasts from the best model. Then, we will have the world's smallest (and fastest) forecasting competition for you to practice your skills via trial by fire, because everyone learns best under pressure! :)

# Set up

We will use a number of libraries for this practical, so please make sure to have them installed and loaded up front. Some, like `tidyverse` aren't absolutely necessary, but we use them here to ease data formatting and for the ggplot dependencies.

```{r, message=FALSE}
library(tidyverse)
library(forecast)
library(Metrics)
```

# Defining the Goal

The first step of forecasting is to define the forecasting goal - why the heck are you about to spend several hours of your life gathering data and modeling it?? Also, once you know that - how are you going to do it?

This process can sometimes take a really long time (even a year or more) for large forecasting projects. It often involves a team of people who have some kind of stake in the forecasting process or results.

In our case, it will just take however much time you need to read the following sentence. The goals of our forecasting exercise are to (1) Practice the forecasting process and (2) Generate 30 day out forecasts for tree greenness using data provided by the NEON Forecasting Challenge and publicly available weather data.

# Getting the Data and Pre-Processing

This data set originally required some cleaning and wrangling. The version you have (`NEONphenologyClean.csv`) is ready to explore and model! Exploring and Visualizing the Data

A simple plot of the time series can go a long way in telling us what kinds of patterns are present. We can already see some strong evidence of seasonality and maybe a slight trend.

```{r, fig.align='center', fig.height=4, fig.width=5}

# Read in the clean data set
phen <- read.csv("data/NEONphenologyClean.csv")

ggplot(data = phen, aes(x = DATE, y = observation))+   
  geom_point() 
```

We can explore some of the predictors we have. There are a lot of them! 
```{r}
names(phen)
```

We can plot some of these and see if it seems like the the patterns in the predictors match up with our main time series. For example, let's say we choose `meanTemp`: 

```{r, fig.align='center', fig.height=4, fig.width=5}

ggplot(data = phen, aes(x = DATE, y = meanTemp))+   
  geom_point() 
```

We can see that temperature seems to share an annual seasonal pattern with our variable of interest. Take some time here and explore some of the other predictors to see how they might match up.

In order to explore our data more easily in R, and also to prepare the data for many of the forecasting functions, we want to formally declare that the data are a time series by turning them into a time series object in R. We do that by using the `ts()` function, which is available in base R. Within the `ts()` function, we need to specify a frequency. The documentation does a good job of explaining what frequency is -- you can take a look here:

```{r}
?ts
```

Given the definition, we can set the frequency here to 365 because the repeating period for trees losing their leaves and greening back up is roughly one year and we have daily observations.

```{r}
phenTS <- ts(phen$observation, frequency = 365) 
```


## Data Exploration

Before we go further, we can get a sense of what sorts of patterns are available in our data. We've seen some of these before, but we also have some short-cut functions that help us visualize our time series data in a quick and dirty way. The most useful of these are the ACF (which we've seen before) and the decomposition plots. 

The *decomposition* plot uses some default moving averages to visualize patterns in the data -- specifically trying to get at the relative amounts of trend, seasonality, and randomness.

```{r}
phenDecomp <- decompose(phenTS) 
plot(phenDecomp)
```

This shows us what that there may be a small trend in the series. It also confirms our intuition that seasonality is present.

We also can look at the ACF to check for the presence of auto-correlation in the time series.

```{r}

acf(phen$observation, lag.max = 365)
```

This shows that there seems to be multiple levels of autocorrelation occurring in the series. On one hand, the series is pretty sticky for nearly 100 days out. On the other hand, the direction of the correlation reverses for certain lags. This also supports the presence of both a trend and seasonal pattern.


# Partitioning the data

When working with time dependent data, data partitioning should not be done randomly. Instead, we will use earlier observations in the data set for training and the remaining observations will be held back as a test set. Generally speaking, it is good to try to match the intended forecasting horizon with the size of the test set. The NEON challenge goal is to forecast at least 30 days out, so we will hold back 30 days of data (i.e. the most recent 30 observations).

```{r}

n <- nrow(phen)

# The length of the data set minus the most recent 30 days
trainN <- n-30
testN <- trainN+1

# Index the earlier rows for training
train <- phen[1:trainN,] 

# Index the later 30 for testing
test <- phen[testN:n,] ## or phen[-1:trainN,]
nrow(test) # Should be 30
```

Notice here that we've got ALL the data still in the testing dataset:
```{r}
summary(test)
```


# Modeling - *Woot Woot 🥳*

We discussed several kinds of models during the lecture. There are many more model types out there that you may encounter, but this set should get you a good start and will provide you with the foundation you need to continue learning on your own should you need to.

In each section, we will fit the model to the training set, generate forecasts to compare to the test set, and generate RMSE and MAE for each model on its test set performance to evaluate each one and compare them in the next section.

## Simple modeling with ARIMA

We know that ARIMA can capture some trends, but is typically less good at seasonal patterns -- and we've got some definite seasonal patterns. But let's give it a go as our default modeling approach first. Remember that this doesn't use any additional predictors, just the original time series, as a time series object:

```{r}
ts.train <- ts(train$observation, frequency = 365)
```



The ARIMA part is fitted via an automated process in the background so we do not have to manually specify $p, d$, and $q$. For more information about this process, see the documentation for the `auto.arima()` function.

Note that this is really really slow....

```{r, cache=TRUE}
fit.arima<-auto.arima(ts.train)
```


```{r}
summary(fit.arima)
```
As always we first check the residuals -- we can look across time, at the ACF, and at the histogram
```{r, fig.align='center', fig.width=5, fig.height=5}
par(mfrow=c(3,1), bty="n", mar = c(4,4,1,1))
plot(fit.arima$residuals)
acf(fit.arima$residuals, main="")
hist(fit.arima$residuals, main="")
```

Not terrible, although we have non-constant variance in different parts of the space. So let's make our first set of forecasts for this model, 30 days out:

```{r}
arima.fc<-forecast(fit.arima, h=30)
```

Then we can make a default plot of the forecast:

```{r}
plot(arima.fc)
```

The forecast isn't too bad. We can also look at the fit plus forecast bounds like in the lectures using base R (note I'm plotting the TESTING data also):

```{r, fig.align='center', fig.width=6, fig.height=5}
plot(as.vector(phen$observation), xlab="day", ylab="pheno", type="l", xlim=c(0, 1560),
     col=1, lwd=2, ylim=c(0.25, 0.55))
lines(as.vector(fit.arima$fitted), col=2, lty=2, lwd=2)
lines(1528+1:30, arima.fc$mean, col="dodgerblue", lwd=2)
lines(1528+1:30, arima.fc$lower[,2], col="dodgerblue", lwd=2, lty=3)
lines(1528+1:30, arima.fc$upper[,2], col="dodgerblue", lwd=2, lty=3)

legend("topleft", legend=c("data", "fit", "FC"), col=c(1:2, "dodgerblue"), lty=c(1,2,1), lwd=2)
```
Pretty good! We can also just zoom in on the last part of the plot.

```{r, fig.align='center', fig.width=6, fig.height=5}
plot(as.vector(phen$observation), xlab="day", ylab="pheno", type="l", xlim=c(1160, 1560),
     col=1, lwd=2, ylim=c(0.3, 0.525))
lines(as.vector(fit.arima$fitted), col=2, lty=2, lwd=2)
lines(1528+1:30, arima.fc$mean, col="dodgerblue", lwd=2)
lines(1528+1:30, arima.fc$lower[,2], col="dodgerblue", lwd=2, lty=3)
lines(1528+1:30, arima.fc$upper[,2], col="dodgerblue", lwd=2, lty=3)

legend("topleft", legend=c("data", "fit", "FC"), col=c(1:2, "dodgerblue"), lty=c(1,2,1), lwd=2)
```

Let's fit one more model then we'll talk about out of sample comparisons...


## Series Decomposition by Loess + ARIMA or Exponential Smoothing

In the lecture portion, we mentioned that one can smooth or fit a model first and then fit an AR or ARIMA model to the residuals. We're going to show an example of this. In this approach the time series is first de-trended and de-seasonalized using something called ***loess*** smoothing. We're going to treat loess as something of a black box, an alternative to the other smoothing methods we've talked about. If you want more details you can take a look here: 


After we've applied the loess smoother, the remaining stationary series is modeled using ARIMA (or another method if we select a different one). 

We will use the `stl` function (also in the `forecast` package) to do our fitting. 
Note, that `stl` (and the related `stlm`) also requires a UNIVARIATE time series object, like we used for the `auto.arima` function above. Further, this function is formally smoothing with lowess then fitting the ARIMA part is in the background using the `auto.arima` function as above. 
 
Below you can see the output for the STL + ARIMA model

```{r}
# There are several options that we can customize in the stl()
#   function, but the one we have to specify is the s.window. 
#   Check out the documentation and play around with some of the
#   other options if you would like.
stl.fit.arima <- stlm(ts.train, 
                      s.window = "periodic",
                      method = "arima")

summary(stl.fit.arima$model)

```


Let's look at the residuals to see if there is anything left unaccounted for.

```{r, fig.align='center', fig.width=5, fig.height=5}
par(mfrow=c(3,1), bty="n", mar = c(4,4,1,1))
plot(stl.fit.arima$residuals)
acf(stl.fit.arima$residuals)
hist(stl.fit.arima$residuals)
```
Again, this looks pretty good, although the variance is non-constant. 


Now that we are comfortable with the model set up, we can generate forecasts. We will do that here for the ARIMA model.

```{r}
# We can generate forecasts with the forecast() function in the
#    forecast package

stl.arima.forecasts <- forecast(stl.fit.arima, h = 30)

# The forecast function gives us point forecasts, as well as
#    prediction intervals
stl.arima.forecasts

# Notice that it is not straightforward to get the point forecasts
#   out - we have to convert it to a data frame first.
str(stl.arima.forecasts)

stl.df <- as.data.frame(stl.arima.forecasts)
```

Let's visualize the forecasts! For the basic models available in the forecast package, we can use the `plot()` function to see a quick time series plot.

```{r}

# Let's look at the forecasts!
plot(stl.arima.forecasts)

```


### Comparing forecasts

Let's compare them to the observed test series values.

```{r}

# First make a data frame with both in there
compare <- data.frame(time = seq(1:30),
                      observed = test$observation,
                      forecast = stl.df$`Point Forecast`)

# What do you think??
ggplot(data = compare, aes(x = time, y = observed))+
  geom_line(color = "blue")+
  geom_point(color = "blue")+
  geom_line(aes(y = forecast), color = "red")+
  geom_point(aes(y = forecast), color = "red")

```

Finally, we can generate some accuracy metrics and save them in a data frame for later use.

```{r}
# Last - let's get the RMSE and MAE out. We can use the Metrics
#    package for this. While we're at it, let's save them to a data
#    set for later.

stl.rmse <- rmse(compare$observed, compare$forecast)
stl.mae <- mae(compare$observed, compare$forecast)

(comparisonDF <- data.frame(model = "stl",
                           RMSE = stl.rmse,
                           MAE = stl.mae))

```

An additional argument in `stlm()` exists that allows for the inclusion of extra predictors into an ARIMA-based model post-decomposition. This function requires the predictors to exist in a separate matrix rather than inside of a data frame, which makes it somewhat challenging to use if you are unfamiliar with using matrices in R.

### Exponential Smoothing

If you want to try out exponential smoothing, you will use `method = "ETS"` and you can specify different model types using `etsmodel = "ANN"` for example. Other potential models can be found in the lecture slides (e.g. "MNN", "AAA", etc.). Note that because we have de-trended and de-seasonalized the series, we do not really need the latter two letters to change from "N".

```{r}
stl.fit.ets <- stlm(ts.train, s.window = "periodic", method = "ets",
                 etsmodel = "ANN")

summary(stl.fit.ets$model)
```


## Linear Regression

#### Slide 29-32

Fortunately, there are several kinds of time series models that accommodate extra predictors in a more intuitive way. The most obvious starting place is linear regression! We have a lot of predictors in our data set to choose from. Feel free to play around with them! For time's sake, I am going to use mean temperature and its one day lag only. This part is probably not too different from what you are used to doing for non-time series data. The only major difference is in what predictors we choose to include to account for autocorrelation, trend, and seasonality.

```{r}

# We will start with just our predictors
lm.fit <- lm(observation ~ meanTemp + meanTempLag1, 
             data = train)
summary(lm.fit)
acf(lm.fit$residuals, 365) # Still a lot of seasonality left

```

Let's add the sine and cosine terms we discussed into the data set to include them in the model. We will need to do this for `train` and `test`. Note that if you were to be forecasting beyond test, you would need to generate these terms for those time points as well.

```{r}

# Add to the training set
train$sinSeason <- sin((2*pi*train$X)/365)
train$cosSeason <- cos((2*pi*train$X)/365)

# Add to the testing set
test$sinSeason <- sin((2*pi*test$X)/365)
test$cosSeason <- cos((2*pi*test$X)/365)

```

Back to business! Let's add in the sine and cosine terms for seasonality.

```{r}
# Add seasonality via sine and cosine terms
lm.fit <- lm(observation ~ sinSeason + cosSeason + meanTemp +
               meanTempLag1, 
             data = train)

summary(lm.fit)
acf(lm.fit$residuals, 365)
```

Finally, we can add in the trend.

```{r}

# Add trend
lm.fit <- lm(observation ~ X + sinSeason + cosSeason + meanTemp +
               meanTempLag1, 
             data = train)

summary(lm.fit)

# Check for leftover autocorrelation in the residuals - still a lot.
#   Stay tuned for how to deal with this! For now we will move 
#   forward.
acf(lm.fit$residuals)

# These models still do assume normal residuals - these look good!
hist(lm.fit$residuals)
```

Next we'll generate forecasts for this model.

```{r}

# We can generate forecasts with the forecast() function in the
#    forecast package
lm.forecasts <- forecast(lm.fit, h = 30, newdata = test)

# The forecast function gives us point forecasts, as well as
#    prediction intervals
lm.forecasts

# Notice that it is not straightforward to get the point forecasts
#   out - we have to convert it to a data frame first.
str(lm.forecasts)

lm.df <- as.data.frame(lm.forecasts)
```

We can visualize the forecasts against the observed values if we add them to our `test` data set.

```{r}

# Add the forecasts to test
test$forecast <- lm.df$`Point Forecast`

# Let's look at the forecasts!

# What do you test??
ggplot(data = compare, aes(x = time, y = observed))+
  geom_line(color = "blue")+
  geom_point(color = "blue")+
  geom_line(aes(y = forecast), color = "red")+
  geom_point(aes(y = forecast), color = "red")
```

Finally we will generate and save the accuracy metrics again.

```{r}

# Last - let's get the RMSE and MAE out. We can use the Metrics
#    package for this. While we're at it, let's save them to a data
#    set for later.

lm.rmse <- rmse(test$observation, test$forecast)
lm.mae <- mae(test$observation, test$forecast)

newrow <- c("lm", lm.rmse, lm.mae)

(comparisonDF <- rbind(comparisonDF, newrow))

```

# Evaluate and Compare Performance

#### Slide 10-12

We have been saving performance metrics for all of our models so far - let's check them out with a graph! Remember - for RMSE and MAE, lower is better. Feel free to take the inverse and look at that if it is confusing to consider this visually.\

```{r}
comparisonDF$RMSE <- as.numeric(comparisonDF$RMSE)
comparisonDF$MAE <- as.numeric(comparisonDF$MAE)

# Which model did the best based on RMSE?
ggplot(data = comparisonDF, aes(x = model, y = RMSE))+
  geom_bar(stat = "identity", fill = "slateblue1")+
  labs(title = "RMSE Comparison")

# Which model did the best based on MAE?
ggplot(data = comparisonDF, aes(x = model, y = MAE))+
  geom_bar(stat = "identity", fill = "plum")+
  labs(title = "MAE Comparison")
```

If we were to want to forecast tree greenness for real, we could try to fine tune the models more to get even better results.

### Next Steps

To generate the real forecasts, we would re-fit the best model using *all* of the data we have. Using the new model parameters, we would then make forecasts out to the desired horizon (in this case 30 days out).

If your chosen model is the ensemble, you would re-fit *all* of the models with *all* of the data and then multiply the forecasts you make by the weights you already generated.

**NOTE:** When you generate the forecasts, you will also have to forecast any predictors you used in your model!! For example, if we used the temperature in our model we would need to either find forecasted temperatures for our horizon or generate other forecasting models to build our own temperature forecasts.
