---
title: "Practical: Time Series Forecasting"
author: "The VectorByte Team (Alicia Arneson S., Virginia Tech)"
format:

  html:

    toc: true

    toc-location: left

    html-math-method: katex

    css: styles.css
editor: visual
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

set.seed(123)
```

# Overview and Instructions

The goals of this practical are to:

1.  Practice setting up data for forecasting

2.  Practice fitting time series models for forecasting

3.  Practice generating forecasts for a given model and evaluating its performance

4.  Learn aggressively.... See the forecasting challenge section below :)

This practical will take you through an example of modeling a terrestrial phenology (how much greenness can be observed from above) data set from NEON and generating forecasts from the best model. Then, we will have the world's smallest (and fastest) forecasting competition for you to practice your skills via trial by fire, because everyone learns best under pressure! :)

# Defining the Goal

#### Slide 4

The first step of forecasting is to define the forecasting goal - why the heck are you about to spend several hours of your life gathering data and modeling it?? Also, once you know that - how are you going to do it?

This process can sometimes take a really long time (even a year or more) for large forecasting projects. It often involves a team of people who have some kind of stake in the forecasting process or results.

In our case, it will just take however much time you need to read the following sentence. The goals of our forecasting exercise are to (1) Practice the forecasting process and (2) Generate 30 day out forecasts for tree greenness using data provided by the NEON Forecasting Challenge and publicly available weather data.

# Getting the Data and Pre-Processing

#### Slide 5 & 7

This data set originally required some cleaning and wrangling. The version you have (NEONphenologyClean.csv) is ready to explore and model! Exploring and Visualizing the Data

#### Slide 6

A simple plot of the time series can go a long way in telling us what kinds of patterns are present. We can already see some strong evidence of seasonality and maybe a slight trend.

```{r, fig.align='center', fig.height=4, fig.width=5}

# Read in the clean data set

phen <- read.csv("NEONphenologyClean.csv")

library(tidyverse)  

ggplot(data = phen, aes(x = DATE, y = observation))+   
  geom_point() 
```

We can explore some of the predictors we have available as well to see if the patterns in the predictors match up. Temperature seems to share a similar seasonal pattern to our variable of interest. Take some time here and explore some of the other predictors to see how they might match up.

```{r, fig.align='center', fig.height=4, fig.width=5}

ggplot(data = phen, aes(x = DATE, y = meanTemp))+   
  geom_point() 
```

Decomposition plots and autocorrelation plots are also useful tools to help us prepare to fit appropriate forecasting models. In order to make a decomposition plot, we have to formally declare that the data are a time series by turning them into a time series object in R. We do that by using the `ts()` function. Several of the more basic time series models also require (or at least prefer) data to be in this format. Within the `ts()` function, we need to specify a frequency. The documentation does a good job of explaining what frequency is:

"Argument `frequency` indicates the sampling frequency of the time series, with the default value `1` indicating one sample in each unit time interval. For example, one could use a value of `7` for `frequency` when the data are sampled daily, and the natural time period is a week, or `12` when the data are sampled monthly and the natural time period is a year."

Given that definition, we can set the frequency here to 365 because the repeating period for trees losing their leaves and greening back up is roughly one year and we have daily observations.

```{r}

phenTS <- ts(phen$observation, frequency = 365) 
```

The decomposition plot shows that there may be a small trend in the series. It also confirms our intuition that seasonality is present.

```{r}

phenDecomp <- decompose(phenTS) 
plot(phenDecomp)
```

The autocorrelation plot shows that there seems to be multiple levels of autocorrelation occurring in the series. On one hand, the series is pretty sticky for nearly 100 days out. On the other hand, the direction of the correlation reverses for certain lags. This also supports the presence of both a trend and seasonal pattern.

```{r}

acf(phen$observation, lag.max = 365)
```

# Partitioning the data

#### Slide 8

Remember, when your goal is forecasting, data partitioning should not be done randomly. The earlier observations in the data set will be used for training and the remaining observations will be held back as a test set. Generally speaking, it is good to try to match the intended forecasting horizon with the size of the test set. The NEON challenge goal is to forecast at least 30 days out, so we will hold back 30 days of data (i.e. the most recent 30 observations).

There are probably packages out there that can do this for you, so feel free to explore other ways of achieving this kind of partitioning!

```{r}

n <- nrow(phen)

# The length of the data set minus the most recent 30 days
trainN <- n-30
testN <- trainN+1

# Index the earlier rows for training
train <- phen[1:trainN,] 

# Index the later 30 for testing
test <- phen[testN:n,]
nrow(test) # Should be 30
```

# Modeling - *Woot Woot 🥳*

We discussed several kinds of models during the lecture. There are many more model types out there that you may encounter, but this set should get you a good start and will provide you with the foundation you need to continue learning on your own should you need to.

In each section, we will fit the model to the training set, generate forecasts to compare to the test set, and generate RMSE and MAE for each model on its test set performance to evaluate each one and compare them in the next section.

## Series Decomposition by Loess + ARIMA or Exponential Smoothing

#### Slide 22-27 & 40-43

In this approach the time series is first de-trended and de-seasonalized using loess smoothing. Then the remaining stationary series is modeled using ARIMA (or another method if we select a different one). The ARIMA part is fitted via an automated process in the background so we do not have to manually specify p, d, and q. For more information about this process, see the documentation for the `auto.arima()` function.

If you want to try out exponential smoothing, you will use `method = "ETS"` and you can specify different model types using `etsmodel = "ANN"` for example. Other potential models can be found in the lecture slides (e.g. "MNN", "AAA", etc.). Note that because we have de-trended and de-seasonalized the series, we do not really need the latter two letters to change from "N".

First, we have to turn our training data into a time series object like we did with the whole series for plotting.

```{r}
library(forecast)

# STL requires a UNIVARIATE time series object
ts.train <- ts(train$observation, frequency = 365)



```

Below you can see the output for the STL + ARIMA and an ETS model

```{r}
# There are several options that we can customize in the stl()
#   function, but the one we have to specify is the s.window. 
#   Check out the documentation and play around with some of the
#   other options if you would like.
stl.fit.arima <- stlm(ts.train, s.window = "periodic", method = "arima")

summary(stl.fit.arima$model)


stl.fit.ets <- stlm(ts.train, s.window = "periodic", method = "ets",
                 etsmodel = "ANN")

summary(stl.fit.ets$model)


```

Let's look at the residuals to see if there is anything left unaccounted for.

```{r}
# Check for leftover autocorrelation in the residuals
acf(stl.fit.arima$residuals)

# These models still do assume normal residuals - these look good!
hist(stl.fit.arima$residuals)
```

Now that we are comfortable with the model set up, we can generate forecasts. We will do that here for the ARIMA model.

```{r}
# We can generate forecasts with the forecast() function in the
#    forecast package

stl.arima.forecasts <- forecast(stl.fit.arima, h = 30)

# The forecast function gives us point forecasts, as well as
#    prediction intervals
stl.arima.forecasts

# Notice that it is not straightforward to get the point forecasts
#   out - we have to convert it to a data frame first.
str(stl.arima.forecasts)

stl.df <- as.data.frame(stl.arima.forecasts)
```

Let's visualize the forecasts! For the basic models available in the forecast package, we can use the `plot()` function to see a quick time series plot.

```{r}

# Let's look at the forecasts!
plot(stl.arima.forecasts)

```

Let's compare them to the observed test series values.

```{r}

# First make a data frame with both in there
compare <- data.frame(time = seq(1:30),
                      observed = test$observation,
                      forecast = stl.df$`Point Forecast`)

# What do you think??
ggplot(data = compare, aes(x = time, y = observed))+
  geom_line(color = "blue")+
  geom_point(color = "blue")+
  geom_line(aes(y = forecast), color = "red")+
  geom_point(aes(y = forecast), color = "red")

```

Finally, we can generate some accuracy metrics and save them in a data frame for later use.

```{r}
# Last - let's get the RMSE and MAE out. We can use the Metrics
#    package for this. While we're at it, let's save them to a data
#    set for later.

library(Metrics)
stl.rmse <- rmse(compare$observed, compare$forecast)
stl.mae <- mae(compare$observed, compare$forecast)

(comparisonDF <- data.frame(model = "stl",
                           RMSE = stl.rmse,
                           MAE = stl.mae))

```

An additional argument in `stlm()` exists that allows for the inclusion of extra predictors into an ARIMA-based model post-decomposition. This function requires the predictors to exist in a separate matrix rather than inside of a data frame, which makes it somewhat challenging to use if you are unfamiliar with using matrices in R.

## Linear Regression

#### Slide 29-32

Fortunately, there are several kinds of time series models that accommodate extra predictors in a more intuitive way. The most obvious starting place is linear regression! We have a lot of predictors in our data set to choose from. Feel free to play around with them! For time's sake, I am going to use mean temperature and its one day lag only. This part is probably not too different from what you are used to doing for non-time series data. The only major difference is in what predictors we choose to include to account for autocorrelation, trend, and seasonality.

```{r}

# We will start with just our predictors
lm.fit <- lm(observation ~ meanTemp + meanTempLag1, 
             data = train)
summary(lm.fit)
acf(lm.fit$residuals, 365) # Still a lot of seasonality left

```

Let's add the sine and cosine terms we discussed into the data set to include them in the model. We will need to do this for `train` and `test`. Note that if you were to be forecasting beyond test, you would need to generate these terms for those time points as well.

```{r}

# Add to the training set
train$sinSeason <- sin((2*pi*train$X)/365)
train$cosSeason <- cos((2*pi*train$X)/365)

# Add to the testing set
test$sinSeason <- sin((2*pi*test$X)/365)
test$cosSeason <- cos((2*pi*test$X)/365)

```

Back to business! Let's add in the sine and cosine terms for seasonality.

```{r}
# Add seasonality via sine and cosine terms
lm.fit <- lm(observation ~ sinSeason + cosSeason + meanTemp +
               meanTempLag1, 
             data = train)

summary(lm.fit)
acf(lm.fit$residuals, 365)
```

Finally, we can add in the trend.

```{r}

# Add trend
lm.fit <- lm(observation ~ X + sinSeason + cosSeason + meanTemp +
               meanTempLag1, 
             data = train)

summary(lm.fit)

# Check for leftover autocorrelation in the residuals - still a lot.
#   Stay tuned for how to deal with this! For now we will move 
#   forward.
acf(lm.fit$residuals)

# These models still do assume normal residuals - these look good!
hist(lm.fit$residuals)
```

Next we'll generate forecasts for this model.

```{r}

# We can generate forecasts with the forecast() function in the
#    forecast package
lm.forecasts <- forecast(lm.fit, h = 30, newdata = test)

# The forecast function gives us point forecasts, as well as
#    prediction intervals
lm.forecasts

# Notice that it is not straightforward to get the point forecasts
#   out - we have to convert it to a data frame first.
str(lm.forecasts)

lm.df <- as.data.frame(lm.forecasts)
```

We can visualize the forecasts against the observed values if we add them to our `test` data set.

```{r}

# Add the forecasts to test
test$forecast <- lm.df$`Point Forecast`

# Let's look at the forecasts!

# What do you test??
ggplot(data = compare, aes(x = time, y = observed))+
  geom_line(color = "blue")+
  geom_point(color = "blue")+
  geom_line(aes(y = forecast), color = "red")+
  geom_point(aes(y = forecast), color = "red")
```

Finally we will generate and save the accuracy metrics again.

```{r}

# Last - let's get the RMSE and MAE out. We can use the Metrics
#    package for this. While we're at it, let's save them to a data
#    set for later.

lm.rmse <- rmse(test$observation, test$forecast)
lm.mae <- mae(test$observation, test$forecast)

newrow <- c("lm", lm.rmse, lm.mae)

(comparisonDF <- rbind(comparisonDF, newrow))

```

# Evaluate and Compare Performance

#### Slide 10-12

We have been saving performance metrics for all of our models so far - let's check them out with a graph! Remember - for RMSE and MAE, lower is better. Feel free to take the inverse and look at that if it is confusing to consider this visually.\

```{r}
comparisonDF$RMSE <- as.numeric(comparisonDF$RMSE)
comparisonDF$MAE <- as.numeric(comparisonDF$MAE)

# Which model did the best based on RMSE?
ggplot(data = comparisonDF, aes(x = model, y = RMSE))+
  geom_bar(stat = "identity", fill = "slateblue1")+
  labs(title = "RMSE Comparison")

# Which model did the best based on MAE?
ggplot(data = comparisonDF, aes(x = model, y = MAE))+
  geom_bar(stat = "identity", fill = "plum")+
  labs(title = "MAE Comparison")
```

If we were to want to forecast tree greenness for real, we could try to fine tune the models more to get even better results.

### Next Steps

To generate the real forecasts, we would re-fit the best model using *all* of the data we have. Using the new model parameters, we would then make forecasts out to the desired horizon (in this case 30 days out).

If your chosen model is the ensemble, you would re-fit *all* of the models with *all* of the data and then multiply the forecasts you make by the weights you already generated.

**NOTE:** When you generate the forecasts, you will also have to forecast any predictors you used in your model!! For example, if we used the temperature in our model we would need to either find forecasted temperatures for our horizon or generate other forecasting models to build our own temperature forecasts.
