---
title: "VectorByte Methods Training: Introduction to Gaussian Processes for Time Dependent Data (Practical Solution)"
author:
  - name: 
      given:  Parul Vijay
      family: Patil
    affiliation: Virginia Tech and VectorByte
  - name: 
      given: Leah R.
      family: Johnson
    url: https://lrjohnson0.github.io/QEDLab/leahJ.html
    affiliation: Virginia Tech and VectorByte
  - name:
      given: Robert B.
      family: Gramacy
    url: https://bobby.gramacy.com/
    affiliation: Virginia Tech
citation: true
date: 2025-06-10
date-format: long
format:
  html:
    toc: true
    toc-location: left
    html-math-method: katex
    css: styles.css
---
# Challenges

Load libraries

```{r, warning=FALSE, message=F}
library(hetGP)
library(tidyverse)
```

The functions we need to borrow:

```{r, warning = F, message = F}

# transforms y
f <- function(x) {
  y <- log(x)
  return(y)
}

# This function back transforms the input argument
fi <- function(y) {
  x <- exp(y)
  return(x)
}

fx.iso_week <- function(datetime){
  startdate <- min(datetime) # identify start week
  x1 <- as.numeric(stringr::str_sub(ISOweek::ISOweek(datetime),7, 8)) # find iso week #
  return(x1)
}

fx.sin <- function(datetime){
  # identify iso week
  x <- as.numeric(stringr::str_sub(ISOweek::ISOweek(datetime),7, 8)) # find iso week 
  # calculate sin value for that week
  x2 <- (sin(2*pi*x/106))^2 
  return(x2)
}

normalize <- function(x, minx, maxx){
  return( (x - minx)/(maxx - minx))
}

```

#### Fit a GP Model for the location "SERC" i.e. `site_number = 7`. 

```{r, warning=FALSE, message=FALSE, dev.args = list(bg = 'transparent'), fig.width= 7, fig.height= 5, fig.align="center", warn.conflicts = FALSE}


target <- readr::read_csv("https://data.ecoforecast.org/neon4cast-targets/ticks/ticks-targets.csv.gz", guess_max = 1e1)

# Visualizing the data
head(target)

# Choose a random site number: Anything between 1-9.
site_number <- 7

# Obtaining site name
site_names <- unique(target$site_id)

# Subsetting all the data at that location
df <- subset(target, target$site_id == site_names[site_number])
df <- df[, c("datetime", "observation")]

cutoff = as.Date('2022-12-31')
df_train <- subset(df, df$datetime <= cutoff)
df_test <- subset(df, df$datetime > cutoff)

# create predictors
X1 <- fx.iso_week(df_train$datetime)
X2 <- fx.sin(df_train$datetime)

# standardize and put into matrix
X1c <- normalize(X1, min(X1), max(X1))
X <- as.matrix(cbind.data.frame(X1c, X2))

# Create a grid from start date in our data set to one year in future (so we forecast for next season)
startdate <- as.Date(min(df$datetime))# identify start week
grid_datetime <- seq.Date(startdate, Sys.Date() + 30, by = 7) # create sequence from 

# Build prediction grid (From 04-2014 to 07-2025)
XXt1 <- fx.iso_week(grid_datetime)
XXt2 <- fx.sin(grid_datetime)

# standardize and put into matrix
XXt1c <- normalize(XXt1, min(XXt1), max(XXt1))
XXt <- as.matrix(cbind.data.frame(XXt1c, XXt2))

# Transform the training response
y_obs <- df_train$observation
y <- f(y_obs + 1)

# Fit a hetGP model. X must be s matrix and nrow(X) should be same as length(y)
het_gpi <- hetGP::mleHetGP(X = X, Z = y)

# Predictions using the base R predict command with a hetGP object and new locationss
het_ppt <- predict(het_gpi, XXt)

# Mean density for predictive locations and Confidence bounds
het_yyt <- het_ppt$mean
het_q1t <- qnorm(0.975, het_ppt$mean, sqrt(het_ppt$sd2 + het_ppt$nugs))
het_q2t <- qnorm(0.025, het_ppt$mean, sqrt(het_ppt$sd2 + het_ppt$nugs)) 

# Back transforming to original scale
het_yy <- fi(het_yyt) - 1
het_q1 <- fi(het_q1t) - 1
het_q2 <- fi(het_q2t) - 1

# Plot Original data
plot(as.Date(df$datetime), df$observation,
       main = paste(site_names[site_number]), col = "black",
       xlab = "Dates" , ylab = "Abundance",
       # xlim = c(as.Date(min(df$datetime)), as.Date(cutoff)),
       ylim = c(min(df_train$observation, het_yy, het_q2), max(df_train$observation, het_yy, het_q1)* 1.2))

# Add testing observations
points(as.Date(df_test$datetime), df_test$observation, col ="black", pch = 19)

# Line to indicate our cutoff point
abline(v = as.Date(cutoff), lwd = 2)

# HetGP Model mean predictions and bounds.
lines(grid_datetime, het_yy, col = 2, lwd = 2)
lines(grid_datetime, het_q1, col = 2, lwd = 1.2, lty = 2)
lines(grid_datetime, het_q2, col = 2, lwd = 1.2, lty =2)

legend("topleft", legend = c("Train Y","Test Y", "HetGP preds"),
         col = c(1, 1, 2), lty = c(NA, NA, 1),
         pch = c(1, 19, NA), cex = 0.5)

```


#### Fit a GP for location "SERC" i.e. `site_number = 7` with X1 = week number

Another predictor that can be used in place of Iso-week (to capture time) is

- $X_1$ Week Number: The week in which the tick density was recorded starting from week 1. Week number does not repeat itself (unlike iso-week).

The function is setup so you have to enter a `datetime`. It identifies the start date and the end date is today's date + 30 days (for short-term future forecasts). 

Next, a sequence of weeks is created and labeled $\{1, 2, \dots, n\}$.

```{r, warning = F, message = F}

# This function tells us the week number given the date from minimum date to maximum date
fx.week <- function(datetime){
  startdate <- min(datetime) # first week of dataset
  currdate <- Sys.Date() + 30 # today's date + 4 weeks from now
  
  # the next few lines identify the mondays of the respective weeks
  start_week <-  ISOweek::ISOweek2date(paste0(ISOweek::ISOweek(startdate), "-1")) 
  curr_week <- ISOweek::ISOweek2date(paste0(ISOweek::ISOweek(currdate), "-1")) 
  
  # all Mondays in dataset
  all_week <- seq.Date(start_week,curr_week, by = 7)
  x1 <- c()
  for (i in 1:length(datetime)){
    # converting each week to monday
    monday <- ISOweek::ISOweek2date(paste0(ISOweek::ISOweek(datetime[i]), "-1"))
    # setting up week number for that week
    x1[i] <- which(monday == all_week)
  }
  return(x1)
}

target$week <- fx.iso_week(target$datetime)
head(target)
```


```{r, warning = F, message = F}
# create predictors
X1 <- fx.week(df_train$datetime)
X2 <- fx.sin(df_train$datetime)

# standardize and put into matrix
X1c <- normalize(X1, min(X1), max(X1))
X <- as.matrix(cbind.data.frame(X1c, X2))

# Create a grid from start date in our data set to one year in future (so we forecast for next season)
startdate <- as.Date(min(df$datetime))# identify start week
grid_datetime <- seq.Date(startdate, Sys.Date() + 30, by = 7) # create sequence from 

# Build prediction grid (From 04-2014 to 07-2025)
XXt1 <- fx.week(grid_datetime)
XXt2 <- fx.sin(grid_datetime)

# standardize and put into matrix
XXt1c <- normalize(XXt1, min(XXt1), max(XXt1))
XXt <- as.matrix(cbind.data.frame(XXt1c, XXt2))

# Transform the training response
y_obs <- df_train$observation
y <- f(y_obs + 1)

# Fit a hetGP model. X must be s matrix and nrow(X) should be same as length(y)
het_gpi <- hetGP::mleHetGP(X = X, Z = y)

# Predictions using the base R predict command with a hetGP object and new locationss
het_ppt <- predict(het_gpi, XXt)

# Mean density for predictive locations and Confidence bounds
het_yyt <- het_ppt$mean
het_q1t <- qnorm(0.975, het_ppt$mean, sqrt(het_ppt$sd2 + het_ppt$nugs))
het_q2t <- qnorm(0.025, het_ppt$mean, sqrt(het_ppt$sd2 + het_ppt$nugs)) 

# Back transforming to original scale
het_yy <- fi(het_yyt) - 1
het_q1 <- fi(het_q1t) - 1
het_q2 <- fi(het_q2t) - 1

# Plot Original data
plot(as.Date(df$datetime), df$observation,
       main = paste(site_names[site_number]), col = "black",
       xlab = "Dates" , ylab = "Abundance",
       # xlim = c(as.Date(min(df$datetime)), as.Date(cutoff)),
       ylim = c(min(df_train$observation, het_yy, het_q2), max(df_train$observation, het_yy, het_q1)* 1.2))

# Add testing observations
points(as.Date(df_test$datetime), df_test$observation, col ="black", pch = 19)

# Line to indicate our cutoff point
abline(v = as.Date(cutoff), lwd = 2)

# HetGP Model mean predictions and bounds.
lines(grid_datetime, het_yy, col = 2, lwd = 2)
lines(grid_datetime, het_q1, col = 2, lwd = 1.2, lty = 2)
lines(grid_datetime, het_q2, col = 2, lwd = 1.2, lty =2)

legend("topleft", legend = c("Train Y","Test Y", "HetGP preds"),
         col = c(1, 1, 2), lty = c(NA, NA, 1),
         pch = c(1, 19, NA), cex = 0.5)

```

#### Fit a GP Model for all the locations with X = (Week number, Iso-week, Sine Predictor)

Hint 1: Write a *function* that can fit a GP and make predictions (keep it simple. Let arguments be the inputs $X$, response $Y$ and predictive set $XX$). 

Hint 2: Write a *for loop* where you subset the data for each location and then call the function to fit the GP.

```{r, warning = F, message = F}
fithgp <- function(X, Y, XX){
  het_gpi <- hetGP::mleHetGP(X = X, Z = Y)
  het_ppt <- predict(het_gpi, XX)
  
  het_yyt <- het_ppt$mean
  het_s2 <- het_ppt$sd2 + het_ppt$nugs
  het_q1t <- qnorm(0.975, het_ppt$mean, sqrt(het_s2))
  het_q2t <- qnorm(0.025, het_ppt$mean, sqrt(het_s2)) 
  return(list(mean = het_yyt, q1 = het_q1t, q2 = het_q2t, s2 = het_s2))
}

site_names <- unique(target$site_id)

par(mfrow = c(1, 1))
for(i in 1:9){
  
  # Subsetting all the data at that location
  dfloc <- subset(target, target$site_id == site_names[i])
  dfloc <- dfloc[, c("datetime", "observation")]

  cutoff = as.Date('2022-12-31')
  df_train <- subset(dfloc, datetime <= cutoff)
  df_test <- subset(dfloc, datetime > cutoff)
  
  X1 <- fx.week(df_train$datetime)
  X2 <- fx.iso_week(df_train$datetime)
  X3 <- fx.sin(df_train$datetime)
  X1c <- normalize(X1, min(X1), max(X1))
  X2c <- normalize(X2, min(X2), max(X2))
  X <- as.matrix(cbind.data.frame(X1c, X2c, X3))

  # Create a grid from start date in our data set to one year in future (so we forecast for   next season)
  startdate <- as.Date(min(dfloc$datetime))# identify start week
  grid_datetime <- seq.Date(startdate, Sys.Date() + 30, by = 7) # create sequence from 
  
  XXt1 <- fx.week(grid_datetime)
  XXt2 <- fx.iso_week(grid_datetime)
  XXt3 <- fx.sin(grid_datetime)

  # standardize and put into matrix
  XXt1c <- normalize(XXt1, min(XXt1), max(XXt1))
  XXt2c <- normalize(XXt2, min(XXt2), max(XXt2))
  XXt <- as.matrix(cbind.data.frame(XXt1c, XXt2c, XXt3))

  # Transform the training response
  y_obs <- df_train$observation
  y <- f(y_obs + 1)
  
  hetfit <- fithgp(X, y, XXt)
  
  # Back transforming to original scale
  het_yy <- fi(hetfit$mean) - 1
  het_q1 <- fi(hetfit$q1) - 1
  het_q2 <- fi(hetfit$q2) - 1
  
  miny <- min(0, dfloc$observation, het_yy, het_q1, het_q2)
  maxy <- max(dfloc$observation, het_yy, het_q1, het_q2) * 1.01
      
  plot(as.Date(dfloc$datetime), dfloc$observation, main = paste(site_names[i]), col =
             "black", xlab = "Dates" , ylab = "Abundance", ylim = c(miny, maxy))
    
  points(as.Date(df_test$datetime), df_test$observation, col ="black", pch = 19)
    
  # Line to indicate our cutoff point
  abline(v = as.Date(cutoff), lwd = 2)
    
    # HetGP Model mean predictions and bounds.
  lines(grid_datetime, het_yy, col = 2, lwd = 2)
  lines(grid_datetime, het_q1, col = 2, lwd = 1.2, lty = 2)
  lines(grid_datetime, het_q2, col = 2, lwd = 1.2, lty =2)
    
  legend("topleft", legend = c("Train Y","Test Y", "HetGP preds"),
             col = c(1, 1, 2), lty = c(NA, NA, 1),
             pch = c(1, 19, NA), cex = 0.5)
}
```

#### Train all sites together

```{r, warning = F, message = F}
hgp <- function(het_gpi, XX){
  het_ppt <- predict(het_gpi, XX)
  het_yyt <- het_ppt$mean
  het_s2 <- het_ppt$sd2 + het_ppt$nugs
  het_q1t <- qnorm(0.975, het_ppt$mean, sqrt(het_s2))
  het_q2t <- qnorm(0.025, het_ppt$mean, sqrt(het_s2)) 
  return(list(mean = het_yyt, q1 = het_q1t, q2 = het_q2t, s2 = het_s2))
}

cutoff = as.Date('2022-12-31')
df_train <- subset(target, datetime <= cutoff)

X1 <- fx.week(df_train$datetime)
X2 <- fx.iso_week(df_train$datetime)
X3 <- fx.sin(df_train$datetime)

X1c <- normalize(X1, min(X1), max(X1))
X2c <- normalize(X2, min(X2), max(X2))
X <- as.matrix(cbind.data.frame(X1c, X2c, X3))

# Transform the training response
y_obs <- df_train$observation
y <- f(y_obs + 1)
  
model <- mleHetGP(X, y)

par(mfrow = c(1, 1))
for(i in 1:9){
  
  # Subsetting all the data at that location
  dfloc <- subset(target, target$site_id == site_names[i])
  dfloc <- dfloc[, c("datetime", "site_id", "observation")]

  df_test <- subset(dfloc, datetime >=cutoff)
  
  # Create a grid from start date in our data set to one year in future (so we forecast for   next season)
  startdate <- as.Date(min(dfloc$datetime))# identify start week
  grid_datetime <- seq.Date(startdate, Sys.Date() + 30, by = 7) # create sequence from 
  
  XXt1 <- fx.week(grid_datetime)
  XXt2 <- fx.iso_week(grid_datetime)
  XXt3 <- fx.sin(grid_datetime)

  # standardize and put into matrix
  XXt1c <- normalize(XXt1, min(XXt1), max(XXt1))
  XXt2c <- normalize(XXt2, min(XXt2), max(XXt2))
  XXt <- as.matrix(cbind.data.frame(XXt1c, XXt2c, XXt3))
  
  hetfit <- hgp(model, XXt)
  
  # Back transforming to original scale
  het_yy <- fi(hetfit$mean) - 1
  het_q1 <- fi(hetfit$q1) - 1
  het_q2 <- fi(hetfit$q2) - 1
  
  miny <- min(0, dfloc$observation, het_yy, het_q1, het_q2)
  maxy <- max(dfloc$observation, het_yy, het_q1, het_q2) * 1.01
      
  plot(as.Date(dfloc$datetime), dfloc$observation, main = paste(site_names[i]), col =
             "black", xlab = "Dates" , ylab = "Abundance", ylim = c(miny, maxy))
    
  points(as.Date(df_test$datetime), df_test$observation, col ="black", pch = 19)
    
      # Line to indicate our cutoff point
  abline(v = as.Date(cutoff), lwd = 2)
    
      # HetGP Model mean predictions and bounds.
  lines(grid_datetime, het_yy, col = 2, lwd = 2)
  lines(grid_datetime, het_q1, col = 2, lwd = 1.2, lty = 2)
  lines(grid_datetime, het_q2, col = 2, lwd = 1.2, lty =2)
    
  legend("topleft", legend = c("Train Y","Test Y", "HetGP preds"),
             col = c(1, 1, 2), lty = c(NA, NA, 1),
             pch = c(1, 19, NA), cex = 0.5)
  
}
```

#### Build altitude predictor for the testing set (*Advanced*).

Here is a snippet of the function to build the predictor.

```{r, warning=FALSE, message=FALSE,}
site_data <- readr::read_csv(paste0("https://raw.githubusercontent.com/eco4cast/neon4cast-targets/","main/NEON_Field_Site_Metadata_20220412.csv")) |> dplyr::filter(ticks == 1)

site_data <- site_data |> dplyr:::select(field_site_id, field_mean_elevation_m)
colnames(site_data) <- c("site_id", "altitude")

# Scaling altitude between [0, 1]
site_data$altitude <- (site_data$altitude - min(site_data$altitude))/(max(site_data$altitude) - min(site_data$altitude))

fx.alt <- function(df, site_df = site_data, test_site = NULL){
  if(!is.null(test_site)) df <- cbind.data.frame(df, site_id = rep(test_site, length(df)))
  df <- df %>% left_join(site_df, by = "site_id")
  return(df$altitude)
}

hgp <- function(het_gpi, XX){
  het_ppt <- predict(het_gpi, XX)
  het_yyt <- het_ppt$mean
  het_s2 <- het_ppt$sd2 + het_ppt$nugs
  het_q1t <- qnorm(0.975, het_ppt$mean, sqrt(het_s2))
  het_q2t <- qnorm(0.025, het_ppt$mean, sqrt(het_s2)) 
  return(list(mean = het_yyt, q1 = het_q1t, q2 = het_q2t, s2 = het_s2))
}

cutoff = as.Date('2022-12-31')
df_train <- subset(target, datetime <= cutoff)

X1 <- fx.week(df_train$datetime)
X2 <- fx.iso_week(df_train$datetime)
X3 <- fx.sin(df_train$datetime)
X4 <- fx.alt(df_train)

X1c <- normalize(X1, min(X1), max(X1))
X2c <- normalize(X2, min(X2), max(X2))
X4c <- normalize(X4, min(X4), max(X4))
X <- as.matrix(cbind.data.frame(X1c, X2c, X3, X4c))

# Transform the training response
y_obs <- df_train$observation
y <- f(y_obs + 1)
  
model <- mleHetGP(X, y)

par(mfrow = c(1, 1))
for(i in 1:9){
  
  # Subsetting all the data at that location
  dfloc <- subset(target, target$site_id == site_names[i])
  dfloc <- dfloc[, c("datetime", "site_id", "observation")]

  df_test <- subset(dfloc, datetime >=cutoff)
  
  # Create a grid from start date in our data set to one year in future (so we forecast for   next season)
  startdate <- as.Date(min(dfloc$datetime))# identify start week
  grid_datetime <- seq.Date(startdate, Sys.Date() + 30, by = 7) # create sequence from 

  XXt1 <- fx.week(grid_datetime)
  XXt2 <- fx.iso_week(grid_datetime)
  XXt3 <- fx.sin(grid_datetime)
  XXt4 <- fx.alt(grid_datetime, test_site = site_names[i])

  # standardize and put into matrix
  XXt1c <- normalize(XXt1, min(XXt1), max(XXt1))
  XXt2c <- normalize(XXt2, min(XXt2), max(XXt2))
  XXt4c <- normalize(XXt4, min(X4), max(X4))
  XXt <- as.matrix(cbind.data.frame(XXt1c, XXt2c, XXt3, XXt4c))
  
  hetfit <- hgp(model, XXt)
  
  # Back transforming to original scale
  het_yy <- fi(hetfit$mean) - 1
  het_q1 <- fi(hetfit$q1) - 1
  het_q2 <- fi(hetfit$q2) - 1
  
  miny <- min(0, dfloc$observation, het_yy, het_q1, het_q2)
  maxy <- max(dfloc$observation, het_yy, het_q1, het_q2) * 1.01
   
  plot(as.Date(dfloc$datetime), dfloc$observation, main = paste(site_names[i]), col =
             "black", xlab = "Dates" , ylab = "Abundance", ylim = c(miny, maxy))
    
  points(as.Date(df_test$datetime), df_test$observation, col ="black", pch = 19)
    
      # Line to indicate our cutoff point
  abline(v = as.Date(cutoff), lwd = 2)
    
      # HetGP Model mean predictions and bounds.
  lines(grid_datetime, het_yy, col = 2, lwd = 2)
  lines(grid_datetime, het_q1, col = 2, lwd = 1.2, lty = 2)
  lines(grid_datetime, het_q2, col = 2, lwd = 1.2, lty =2)
    
  legend("topleft", legend = c("Train Y","Test Y", "HetGP preds"),
             col = c(1, 1, 2), lty = c(NA, NA, 1),
             pch = c(1, 19, NA), cex = 0.5)
  
}
```
